---
title: 'Lab 9: Cluster and Cox processes'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session is concerned with summary statistics for interpoint
spacing and distances.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script09.R) (right click and save).

```{r include=FALSE}
library(spatstat)
```

### Exercise 1

The command `rThomas` generates simulated realisations of
the Thomas model (‘modified Thomas cluster process’).

1.  Read the help file.

See `help("rThomas")`.

2.  Type `plot(rThomas(10, 0.05, 8))` a few times, and
    interpret the results.

```{r, results="hide", echo=-1}
par(mar=rep(0,4))
replicate(3, plot(rThomas(10, 0.05, 8), main = ""))
```

A clustered process -- on average 10 clusters with 8 points with a standard deviation of 0.05 from their parent (i.e. approximately at distance 0.05 from their parent on average.

3.  Experiment with the arguments of `rThomas` to obtain
    point patterns that
    1.  consist of a few, well-separated, very tight clusters of
        points;
    2.  look similar to realisations of a uniform Poisson process.

We get few clusters by reducing the intensity of the parent process (first argument).
Tightly and separated clusters are obtained by reducing the standard deviation (second argument).

```{r, echo=-1}
par(mar=rep(0,4))
plot(rThomas(5, 0.01, 8), main = "")
```

We simulate from a Poisson process with the same intensity as the parent process of the Thomas process:

```{r, echo=-1}
par(mar=rep(0,4))
plot(rpoispp(5), main = "")
```


### Exercise 2

1.  Read the help file for `kppm`.

See `help("kppm")`.

2.  Fit the Thomas model to the `redwood` data by the
    method of minimum contrast:
    ```{r, eval=FALSE}
    fit <- kppm(redwood ~ 1, clusters="Thomas")
    fit
    plot(fit)
    ```

From the documentation, the minmum contrast fitting procedure is default. Hence, we need not specify it.

```{r}
fit <- kppm(redwood ~ 1, clusters = "Thomas")
fit
plot(fit, main = "") 
```

3.  Read off the parameters of the fitted model, and generate a
    simulated realisation of the fitted model using `rThomas`.

From the previous output, we can read off the parameters to do the simulation:

```{r, echo=-1}
par(mar=rep(0,4))
rt2 <- rThomas(kappa = 23.55, scale = 0.04, mu = 2.63)
plot(rt2, main = "")
```

4.  Type `plot(simulate(fit))` to generate a simulated realisation
    of the fitted model automatically.

OK, let try that alternative:

```{r, echo=-1}
par(mar=rep(0,4))
plot(simulate(fit), main = "", main.panel = "")
```

5.  Try the command
    ```{r}
    fit2 <- kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=10, scale=0.1))
    ```
    and briefly explore the fitting algorithm’s sensitivity to the
    initial guesses at the parameter values `kappa` and `scale`.

For "large" kappa (parent intensity) and "small" scale (standard deviation), the 
algorithm seems quite robust:
```{r}
kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=10, scale=0.1))
kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=100, scale=0.01))
```

However, for a very small parent intensity (kappa) and large offspring scale the fit is poor. This basically corresponds a Poisson point process.

```{r}
kppm(redwood ~ 1, clusters="Thomas", startpar=c(kappa=0.1, scale=10))
```

6.  Generate and plot several simulated realisations of the fitted
    model, to assess whether it is plausible.

```{r, echo=-1}
par(mar=rep(0,4))
plot(simulate(fit, nsim = 12), main = "", main.panel = "")
plot(redwood)
```

The actual data do not look too different from the simulated.

7.  Extract and plot the fitted pair correlation function by
    ```{r, eval = FALSE}
    pcffit <- pcfmodel(fit)
    plot(pcffit, xlim = c(0, 0.3))
    ```

OK, let's try that:

```{r}
pcffit <- pcfmodel(fit)
plot(pcffit, xlim = c(0, 0.3), main = "pair correlation")
```

8.  Type `plot(envelope(fit, Lest, nsim=39))` to generate simulation
    envelopes of the $L$ function from this fitted model. Do they
    suggest the model is plausible?

```{r, results='hide'}
plot(envelope(fit, Lest, nsim = 39, global = TRUE))
```

Yes, the model seems plausible and it does not devivated from the envelope.


### Exercise 3

1.  Fit a Matern cluster process to the `redwood` data.

We fit the Matern cluster process by specifying the `clusters` argument to be `MatClust`.

```{r}
mfit <- kppm(redwood ~ 1, clusters = "MatClust")
```

2.  Use `vcov` to estimate the covariance matrix of the parameter estimates.

The variance (covariance matrix) is computed straightforwardly:

```{r}
vcov(mfit)
```

3.  Compare with the covariance matrix obtained when fitting a homogeneous Poisson model.

```{r}
vcov(ppm(redwood ~ 1))
```

As can be seen, the variance of the intensity estimate is quite a bit larger in the Matern model. This comes naturally by the doubly stochasic of the Matern model.

