---
title: 'Lab 8: Spacing and distances'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session is concerned with summary statistics for
spacings and interpoint distances.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script08.R) (right click and save).

```{r include=FALSE}
library(spatstat)
```

### Exercise 1

For the `swedishpines` data:

1.  Calculate the estimate of the nearest neighbour distance
    distribution function $G$ using `Gest`.

```{r}
G <- Gest(swedishpines)
```

2.  Plot $\hat G(r)$ against $r$

```{r, results='hide'}
plot(G, cbind(km, rs, han) ~ r, main = "Nearest neighbor distance distribution")
```

3.  Plot $\hat G(r)$ against the theoretical (Poisson) value
    $G_{\mbox{pois}}(r) = 1 - \exp(-\lambda \pi r^2)$.

E.g.

```{r, results='hide'}
plot(G, . ~ theo, main = "Nearest neighbor distribution")
```

4.  Define Fisher’s variance-stabilising transformation for c.d.f.’s
    by
    ```{r, eval = FALSE}
    Phi <- function(x) asin(sqrt(x))
    ```
    Plot the $G$ function using the formula `Phi(.) ~ Phi(theo)`
    and interpret it.
    
```{r, eval = FALSE}
Phi <- function(x) asin(sqrt(x))
plot(G, Phi(.) ~ Phi(theo), main = "Nearest neighbor distribution")
```

Without envelopes it's hard to tell if the functions deviate from the identity.

### Exercise 2

For the `swedishpines` data:

1.  Calculate the estimate of the nearest neighbour distance
    distribution function $F$ using `Fest`.

```{r}
Fhat <- Fest(swedishpines)
```

2.  Plot $\hat F(r)$ against $r$

```{r, results='hide'}
plot(Fhat, main = "Empty Space function distribution")
```

3.  Plot $\hat F(r)$ against the theoretical (Poisson) value
    $F_{\mbox{pois}}(r) = 1 - \exp(-\lambda \pi r^2)$.

```{r, results='hide'}
plot(Fhat, . ~ theo, main = "")
```

4.  Define Fisher’s variance-stabilising transformation for c.d.f.’s
    by
    ```{r, eval = FALSE}
    Phi <- function(x) asin(sqrt(x))
    ```
    Plot the $F$ function using the formula `Phi(.) ~ Phi(theo)`
    and interpret it.

```{r, results='hide'}
Phi <- function(x) asin(sqrt(x))
plot(Fhat, Phi(.) ~ Phi(theo), main = "")
```

Again, without envelopes it's hard to conclude whether there is any difference.
